---
title: "ST 558 Fall 2026 Project One"
author: "Cody Ashby"
format: html
editor: visual
---

This project will involve querying a census API, getting specific data sets from it, and writing customized functions that summarize and plot certain parts of the data.

# Part One: Data Processing

Since we'll be looking at a small handful of `json` files, we'll need to install (if necessary) a couple of packages, namely the `httr` and `jsonlite` packages.

```{r}
# Install if necessary
install.packages("httr")

library(httr)
library(jsonlite)
library(tidyverse)
```

## Parsing Variable Names

Now, a function will be created to retrieve the URL, parse the JSON, and pull out a list of variables with their values.

```{r}
#Collection of variables for 2021 and 2022 data
vars_list_one=c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")
#Collection of variables for 2023 and 2024 data
vars_list_two=c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")

get_PUMS_vars=function(year){
  API_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums/variables.json"))
  parsed_url<-fromJSON(rawToChar(API_url$content))
  parsed_vars=parsed_url$variables
  vars_list_one=c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")
  vars_list_two=c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")
  if(year %in% c(2021,2022)){
    parsed_vars_select=parsed_vars[vars_list_one]
  }
  else if(year %in% c(2023,2024)){
    parsed_vars_select=parsed_vars[vars_list_two]
  }
  else{
    return("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  return(parsed_vars_select)
}

PUMS_vars_2021=get_PUMS_vars(2021)
PUMS_vars_2022=get_PUMS_vars(2022)
PUMS_vars_2023=get_PUMS_vars(2023)
PUMS_vars_2024=get_PUMS_vars(2024)

all_PUMS_vars=c(PUMS_vars_2021,PUMS_vars_2022,PUMS_vars_2023,PUMS_vars_2024)

saveRDS(all_PUMS_vars,"all_PUMS_vars.rds")
```

The variables of interest are AGEP, GASP, GRPIP, JWAP, JWMNP, PWGTP, FER, HHL, SCL, SCHL, SEX, ST (2021 and 2022) or STATE (2023 and 2024), REGION, and DIVISION.

Four lists will be required for each of the years 2021, 2022, 2023, and 2024 and will be saved in an `.rds` file under the `data` folder in the repository.

## Parsing API Data

The ultimate goal for this section is to write a function that queries the PUMS API and returns a nicely formatted tibble. To kick things off, we'll use a helper function as defined below:

```{r}
get_PUMS_data=function(year){
  if(year %in% c(2023,2024)){
    test_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,PWGTP,FER,HHL,SCH,SCHL,SEX,STATE,REGION,DIVISION&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else if(year %in% c(2021,2022)){
    test_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,PWGTP,FER,HHL,SCH,SCHL,SEX,ST,REGION,DIVISION&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else{
    stop("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  parsed_url=fromJSON(rawToChar(test_url$content))
  colnames(parsed_url)=as.character(parsed_url[1,])
  formatted_data=as_tibble(parsed_url[-1,])
  return(formatted_data)
}
```

Let's try this out! In fact, we'll use the object we get later.

```{r}
#saved as an object for later use
PUMS_2022=get_PUMS_data(2022)
PUMS_2022
```

Next, we'll use elements of that helper function to create a new function that allows a user to query the API, select certain variables from that data set, and parse it.

```{r}
select_PUMS_data=function(year=2024,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level="ST",sublevel=NULL){
  if(year %in% c(2021,2022,2023,2024)){
  API_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=",numvar1,",",other_numvars,",",catvar1,",",other_catvars,",",level,"&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else{
    stop("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  numvar1 <- as.numeric(numvar1)
  if(other_numvars %in% c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP")){
    other_numvars <- as.numeric(other_numvars)
    if(other_numvars %in% c("JWAP","JWDP")){
      other_numvars <- factor(other_numvars,levels=all_PUMS_vars$other_numvars$values$item)
    }
  }
  catvar1=factor(catvar1,levels=all_PUMS_vars$catvar1$values$item)
  if(other_catvars %in% c("FER","HHL","SCH","SCHL")){
    other_catvars <- factor(other_catvars,levels=all_PUMS_vars$other_catvars$values$item)
  }
  if(level=="REGION"){
    sublevel=="South"
  }
  if(level=="DIVISION"){
    sublevel=="South Atlantic (South Region)"
  }
  if(level %in% c("ST","STATE")){
    sublevel=="North Carolina/NC"
  }
  else{
    stop("Not a valid variable!")
  }
  if(level=="All"){
    stop("Only ONE geography level is allowed! Region, Division, or State.")
  }
  parsed_url=fromJSON(rawToChar(API_url$content))
  colnames(parsed_url)=as.character(parsed_url[1,])
  formatted_data=as_tibble(parsed_url[-1,])
  class(formatted_data) <- c("census",class(formatted_data))
  return(formatted_data)
}
```

Let's try this out!

```{r}
#saved as an object for later use
PUMS_2022_subset=get_PUMS_data(2022,"PWGTP","AGEP","SEX","SCH")
PUMS_2022_subset
```

## Combining Years

Here, we'll use this perplexing function multiple times to create a tibble that covers multiple years while still subsetting the variables we need.

```{r}
select_multi_PUMS_data<-function(year1,year2,year3=NULL,year4=NULL,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level="ST",sublevel=NULL){
  if(any(c(year1,year2,year3,year4) %in% c(2023,2024))){
    level=="STATE"
  }
  PUMS_one=select_PUMS_data(year1,numvar1,other_numvars,catvar1,other_catvars,level,sublevel)
  PUMS_two=select_PUMS_data(year2,numvar1,other_numvars,catvar1,other_catvars,level,sublevel)
  if(!is.null(year3) | !is.null(year4)){
    PUMS_three=select_PUMS_data(year3,numvar1,other_numvars,catvar1,other_catvars,level,sublevel)
    PUMS_four=select_PUMS_data(year4,numvar1,other_numvars,catvar1,other_catvars,level,sublevel)
    }
  bind_rows(PUMS_one,PUMS_two)
}
```

Let's try this function for the years 2021 and 2022.

```{r}
select_multi_PUMS_data(2021,2022,"PWGTP","AGEP","SEX","SCHL")
```

# Part Two: Writing a Generic Function for Summarizing

## Numerical Summaries

Below is a function that takes a tibble and produces means and standard deviations for numeric variables (ignoring time variables) and counts for categorical variables.

```{r}
number_sum <- function(tibble,numvar,catvar,...){
  weights=tibble$PWGTP
  mean=sum((tibble$numvar)*weights)/sum(weights)
  sd=sqrt(sum((tibble$numvar)^2*weights)/sum(weights)-mean(tibble$numvar)^2)
  count=table(tibble$catvar)
  return(list(mean,sd,count))
}
```

Let's see how the ages and school enrollments are distributed here.

```{r}
number_sum(PUMS_2022,"AGEP","SCH")
```

## Graphical Summaries

To write a function that produces a boxplot of one categorical and one numeric variable from a tibble. Elements from `ggplot2` will be used here.

```{r}
graph_sum <- function(tibble,catvar,numvar){
  ggplot(data=tibble,aes(x=as.factor(catvar),y=numvar,weight=PWGTP)) +
    geom_boxplot() + 
    labs(x=get(catvar),y=get(numvar)) +
    theme_minimal()
}
```

Let's visualize the numeric summary from above.

```{r}
graph_sum(PUMS_2022,"SCH","AGEP")
```

# Results

No matter what year we choose or the subset of variables we're looking at, these are incredibly large data sets!
