---
title: "ST 558 Fall 2026 Project One"
author: "Cody Ashby"
format: html
editor: visual
---

This project will involve querying a census API, getting specific data sets from it, and writing customized functions that summarize and plot certain parts of the data.

```{r}
# URLs to use:

#Info about PUMS API
https://www.census.gov/data/developers/data-sets/census-microdata-api.html

#Variables for 2024  
https://api.census.gov/data/2024/acs/acs1/pums/variables.html

#Allows accuess to certain subsets of data
https://data.census.gov/app/mdat/
  
#Data sets to read in and parse  
https://api.census.gov/data/2022/acs/acs1/pums/variables.json

https://api.census.gov/data/2023/acs/acs1/pums/variables.json

https://api.census.gov/data/2021/acs/acs1/pums/variables.json

https://api.census.gov/data/2024/acs/acs1/pums/variables.json

#Use this to specify geography level
https://api.census.gov/data/2024/acs/acs1/pums/geography.html


```


# Part One: Data Processing

Since we'll be looking at a small handful of `json` files, we'll need to install (if necessary) a couple of packages, namely the `httr` and `jsonlite` packages.

```{r}
# Install if necessary
install.packages("httr")

library(httr)
library(jsonlite)
library(tidyverse)
```



## Parsing Variable Names

Now, a function will be created to retrieve the URL, parse the JSON, and pull out a list of variables with their values. 

```{r}
#Collection of variables for 2021 and 2022 data
vars_list_one=c("AGEP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")
#Collection of variables for 2023 and 2024 data
vars_list_two=c("AGEP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")

get_PUMS_vars=function(year){
  API_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums/variables.json"))
  parsed_url<-fromJSON(rawToChar(API_url$content))
  parsed_vars=parsed_url$variables
  vars_list_one=c("AGEP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")
  vars_list_two=c("AGEP","GRPIP","JWAP","JWDP","JWMNP","PWGTP",
                "FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")
  if(year %in% c(2021,2022)){
    parsed_vars_select=parsed_vars[vars_list_one]
  }
  else if(year %in% c(2023,2024)){
    parsed_vars_select=parsed_vars[vars_list_two]
  }
  else{
    return("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  return(parsed_vars_select)
}

PUMS_vars_2021=get_PUMS_vars(2021)
PUMS_vars_2022=get_PUMS_vars(2022)
PUMS_vars_2023=get_PUMS_vars(2023)
PUMS_vars_2024=get_PUMS_vars(2024)

all_PUMS_vars=c(PUMS_vars_2021,PUMS_vars_2022,PUMS_vars_2023,PUMS_vars_2024)

saveRDS(all_PUMS_vars,"all_PUMS_vars.rds")
```

The variables of interest are AGEP, GASP, GRPIP, JWAP, JWMNP, PWGTP, FER, HHL, SCL, SCHL, SEX, ST (2021 and 2022) or STATE (2023 and 2024), REGION, and DIVISION.

Four lists will be required for each of the years 2021, 2022, 2023, and 2024 and will be saved in an `.rds` file under the `data` folder in the repository.

## Parsing API Data

The ultimate goal for this section is to write a function that queries the PUMS API and returns a nicely formatted tibble. To kick things off, we'll use a helper function as defined below:

```{r}
get_PUMS_data=function(year){
  if(year %in% c(2023,2024)){
    test_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,PWGTP,FER,HHL,SCH,SCHL,SEX,STATE,REGION,DIVISION&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else if(year %in% c(2021,2022)){
    test_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,PWGTP,FER,HHL,SCH,SCHL,SEX,ST,REGION,DIVISION&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else{
    stop("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  parsed_url=fromJSON(rawToChar(test_url$content))
  colnames(parsed_url)=as.character(parsed_url[1,])
  formatted_data=as_tibble(parsed_url[-1,])
  return(formatted_data)
}
```

Next, we'll use elements of that helper function to create a new function that allows a user to query the API, select certain variables from that data set, and parse it.

```{r}
select_PUMS_data=function(year=2024,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level=contains("ST"),sublevel=NULL){
  all_PUMS_vars=readRDS(all_PUMS_vars.rds)
  if(year %in% c(2021,2022,2023,2024)){
  API_url=GET(paste0("https://api.census.gov/data/",year,"/acs/acs1/pums?get=",numvar1,other_numvars,catvar1,other_catvars,level,"&key=af787134d167e33ea1a00ea6f50f6961418ab20b"))
  }
  else{
    stop("Not a valid year! Use 2021, 2022, 2023, or 2024.")
  }
  if(c(numvar1,other_numvars) %in% c("PWGTP","AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP")){
    c(numvar1,other_numvars) <- as.numeric(c(numvar1,other_numvars))
    if(other_numvars %in% c("JWAP","JWDP")){
      other_numvars <- factor(other_numvars,labels=other_numvars$values$item,levels=other_numvars$values$item)
    }
  }
  if(c(catvar1,other_catvars) %in% c("SEX","FER","HHL","SCH","SCHL")){
    c(catvar1,other_catvars) <- factor(c(catvar1,other_catvars),labels=c(catvar1,other_catvars)$values$item,levels=c(catvar1,other_catvars)$values$item)
  }
  if(level=="REGION"){
    sublevel=="South"
  }
  if(level=="DIVISION"){
    sublevel=="South Atlantic (South Region)"
  }
  if(level==contains("ST")){
    sublevel=="North Carolina/NC"
  }
  if (!(c(numvar1,other_numvars,catvar1,other_catvars,level) %in% all_PUMS_vars)){
    stop("Not a valid variable!")
  }
  if(level=="All"){
    stop("Only ONE geography level is allowed! Region, Division, or State.")
  }
  parsed_url=fromJSON(rawToChar(API_url$content))
  colnames(parsed_url)=as.character(parsed_url[1,])
  formatted_data=as_tibble(parsed_url[-1,])
  class(formatted_data) <- c("census",class(formatted_data))
  return(formatted_data)
}

# Wrap the helper function into a function that also parses the data by querying the API and allow the user to alter the following: year (default is 2024), numeric variables (AGEP and PWGTP as default; options should include AGEP, GASP, GRPIP, JWAP, JWDP, and JWMNP; also, turn these variables into numeric values or time values via lubridate; for any missing time values, use "NA"), categorical variables (SEX as default; options should include FER, HHL, SCH, SCHL and SEX; turn these variables into factors with appropriate levels - may need to use the rds file).

# If user specifies variable not included above, or does not use required variables, retuen an error.
# Also, specify geography level: region, division, or state (last as default). Do not use "All". Only one is allowed; if more than one or "all" is input, return an error.

#Specify specific region, division, or state for subsetting (specify default level for each; use allowable values from rds file; include option for returning "all" values)

#Before returning tibble, include this line of code:

#class(tbd_tibble) <- c("census",class(tbd_tibble))


```

## Combining Years

Here, we'll use this perplexing function multiple times to create a tibble that covers multiple years while still subsetting the variables we need.

Write function that allows specification of multiple years of data along with other options from previous part. Call single year function as many times as needed, then combine into one final tibble (include year variable in final tibble).

```{r}
select_multi_PUMS_data<-function(year1,year2,year3=NULL,year4=NULL){
  PUMS_one=select_PUMS_data(year1,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level=contains("ST"),sublevel=NULL)
  PUMS_two=select_PUMS_data(year2,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level=contains("ST"),sublevel=NULL)
  if(!is.null(year3) | !is.null(year4)){
      PUMS_three=select_PUMS_data(year3,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level=contains("ST"),sublevel=NULL)
      PUMS_four=select_PUMS_data(year4,numvar1="PWGTP",other_numvars="AGEP",catvar1="SEX",other_catvars=NULL,level=contains("ST"),sublevel=NULL)
    }
  bind_rows(PUMS_one,PUMS_two,PUMS_three,PUMS_four)
}
```

# Part Two: Writing a Generic Function for Summarizing

## Generic Functions

Not sure if we're actually supposed to do anything here? If so, then...

```{r}
summary.census <- function(tibble){
  summary(tibble)
}
```

## Numeric Summaries

Write a function that produces means and stdevs for numeric variables (ignoring time variables) and counts for categorical variables.

```{r}
number_sum <- function(tibble,numvar,...,catvar,...){
  weights=tibble$PWGTP
  mean=sum((tibble$numvar)*weights)/sum(weights)
  sd=sqrt(sum((tibble$numvar)^2*weights)/sum(weights)-mean(tibble$numvar)^2)
  count=table(tibble$catvar)
  return(list(mean,sd,count))
}
```

## Graphical Summaries

Write generic `plot()` function that requires one categorical variable (including year) and one numeric variable; this should return a boxplot of the numeric variable at the levels of the categorical variable. Use `weight=PWGTP` in the `aes()` portion so that each row is treated appropriately.

```{r}
graph_sum <- function(catvar,numvar){
  ggplot(data=tibble,aes(x=as.factor(catvar),y=numvar,weight=PWGTP)) +
    geom_boxplot() + 
    labs(x=get(catvar),y=get(numvar)) +
    theme_minimal()
}
```

# Results
